Implementing Distributed Systems Directly with I/O Automata in Ioa++

### 2011-09-20 papers/paper1/component_model.tex
Dr. Gill's comments on the component model.

Modularity, well-defined interfaces, and composition are essential properties of reusable software.
A unit of software that exhibits these properties is called a \emph{component}~\cite{szyperski2002component}.
Modularity implies that a component can be deployed independently of other components.
Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions with the component.
Composition allows a group of interacting components to be understood as a cohesive unit.

In this section we describe how I/O automata offer a natural realization of components and interfaces, and provide natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that is consumed by zero or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then applying the effect if the precondition is true.
In general the scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
However, it lacks certain features needed to develop real components.
This subsection introduces these features and lays the foundation for adding the dynamics discussed in section~\ref{dynamics}.

\paragraph{Recursive Encapsulation.}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up and top-down design strategies.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children and an anonymous \emph{kernel automaton}.
The kernel automaton allows a parent automaton to have its own state variables and actions.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
The name of a named automata appears after its type, e.g., Converter C.
The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

\paragraph{Explicit Composition.}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and is a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.

For the event detection system illustrated in Figure~\ref{sys_model}, actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have no suffix, e.g., \emph{process1}.
The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.
The map of bindings for the event detection system is $\{ (out!(s), C.in?(s)), (C.out!(t), D.in?(t)), (D.detect!, detected?) \}$.

\paragraph{Binding rules.}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), C.in?(s))$ agree on type $s$, $(C.out!(t), D.in?(t))$ agree on type $t$, and $(D.detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
Consequently, every binding has exactly one owner.
For a graphical representation such as Figure~\ref{sys_model}, this rule means that an input is the sink of at most one arrow (binding).
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
Graphically, arrows originating at the same output action terminate in different components.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph{Concurrent execution.}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound, i.e., composed.
Two actions can be executed concurrently if their respective sets of implied automata and therefore state variables are disjoint.
The implied automata sets for each local action in figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.

These observations create some interesting opportunities for designing and analyzing concurrent software.
Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
A high degree of fan-out (an output action being bound to many input actions) creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
However, the input effects can be applied in parallel since the state of each automaton is independent.
A high degree of fan-in (an automaton has many bound input actions) also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
A scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
Compositions of automata can be statically composed to take advantage of in-lining.
An individual automaton can be factored into a number of child automata to increase parallelism.

\paragraph{Parameters.}
Actions in the I/O automata model can be defined using parameters.
Parameters are useful for situations requiring fan-in or session semantics.
For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

\paragraph{Action types.}
The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
unvalued unparameterized output (uv-up output),
unvalued parameterized output (uv-p output),
valued unparameterized output (v-up output),
valued parameterized output (v-p output),
unvalued unparameterized input (uv-up input),
unvalued parameterized input (uv-p input),
valued unparameterized input (v-up input),
valued parameterized input (v-p input),
unparameterized internal (up internal), and
parameterized internal (p internal).

@@

\paragraph{Example.}
Figure~\ref{example1} depicts a remote sensor systems as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton $Z$ and three Channel automata $A$, $B$, and $C$ representing three network connections.
The Control automaton $Z$ consists of the Sensor automaton $S$, the Filter automaton $F$, and the Statistics automaton $T$.
The Control automaton $C$ starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter $F$ and Statistics component $T$ via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton $F$ passes the filtered sample to the Control automaton $Z$ via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton $T$ calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton $Z$ using the bindings $(Z.request!, T.request?)$ and $(T.response!(u), Z.response?(u))$.
The Control automaton $Z$ contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata $A$, $B$, and $C$.
The parameters 0, 1, and 2, are used to route messages to Channel automata $A$, $B$, and $C$ respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

The I/O automata model assumes that systems are composed of a finite or infinitely countable number of automata~\cite{lynch1996distributed}.
Assuming an infinitely countable number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Consequently, we require the ability to deal with a dynamic set of automata and interactions.

\paragraph{System actions.}
The four indivisible operations for managing a dynamic set of automata are create, bind, unbind, and destroy.
Create, bind, unbind, and destroy are collectively called \emph{system actions}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in section~\ref{practical} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exist so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy.

@@

\paragraph{The system automaton.}
Let the set of automata $A$ and set of bindings $B$ be managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata exist are selected and uses $B$ to ensure that the appropriate set of inputs actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in section~\ref{practical} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

@@

To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
To do this, we conjunct the precondition of an output action with a predicate over the set of bindings $B$ to ensure that the output action is not executed until its set of implied automata is correct.
Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as previously discussed.
Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
A static system can be implemented by adding binding predicates and logic to create the required constellation of automata.

>>>>

Modularity, well-defined interfaces, and composition are essential properties of reusable software.
A unit of software that exhibits all of these properties is called a \emph{component}~\cite{szyperski2002component}.
Modularity implies that a component can be deployed independently of other components.
Well-defined interfaces and interactions allow components to expose their functionality in a regular way that facilitates reasoning about different interactions among them.
Composition allows a group of interacting components to be understood as a cohesive unit.

In this section we describe how the I/O automata model offers a natural realization of components and interfaces, and provides natural support for concurrency and composition.
The Input/Output (I/O) automata model was developed by Lynch~\cite{lynch1996distributed} to model reactive concurrent and asynchronous systems.
An I/O automaton consists of state variables and a set of atomic actions that manipulate the state variables.
An action is only permitted to manipulate the state variables of the automaton to which it belongs.
An action is either an internal action, an output action, or an input action.
An internal action changes the state of the automaton.
An output action changes the state of the automaton and produces a signal or value that may be consumed by one or more input actions.
An input action changes the state of the automaton when it receives a signal or value produced by an output action.
\emph{Unvalued} output and input actions produce and consume signals respectively.
\emph{Valued} output and input actions produce and consume values respectively.
The set of actions associated with an automaton is known as the automaton's \emph{signature}~\cite{lynch1996distributed}.

@@

Execution in the I/O automata model consists of non-deterministically and repeatedly selecting a local action and then if the precondition is true applying the effect.
The scheduler is assumed to be fair, meaning that a local action is guaranteed to be selected (but not necessarily executed) infinitely often.
Executing internal actions and output actions that are not composed with any input actions is straightforward: in one atomic step, the precondition is evaluated and the effect is applied if the precondition is true.
Executing a composed output action consists of atomically evaluating the precondition and applying the effect of the output action and all composed input actions,  subject to the precondition using the value produced by the output action.
The atomic relationship between composed output actions and input actions means that multiple automata may change state atomically when executing an output action.
The guarantee of atomicity allows a developer to relate the state of one automaton to another, resulting in well-defined interactions that facilitate composition.

@@

As a mathematical abstraction the I/O automata model is appropriate for modeling reactive asynchronous and concurrent systems.
However, it lacks certain features needed to develop real components.
This subsection introduces these features and lays the foundation for adding the dynamics discussed in Section~\ref{dynamics}.

\paragraph{Recursive Encapsulation.}
Component frameworks that support recursive encapsulation allow components to contain other components.
Recursive encapsulation is at the heart of both bottom-up design strategies where smaller components are composed to produce larger ones, and top-down design strategies in which larger components are decomposed into smaller ones.
The contained component is the \emph{child} while the containing component is the \emph{parent}.
The resulting hierarchy helps developers organize programs by separating concerns and facilitates the reuse of existing components.
A distinction is made between the \emph{name} given to an instance and the \emph{type} of a child component to allow a parent component to contain multiple children of the same type.

Applying recursive encapsulation to I/O automata means that an I/O automaton may contain and use other I/O automata.
Recursive encapsulation implies that a parent automaton is the result of composing its children.
This definition excludes a parent from having any state or actions of its own.
Since it is often convenient to associate state and actions with a parent, we allow a parent to be the composition of its children and an anonymous child that contains the state and actions attributed to the parent.
The \emph{root automaton} is the automaton at the top of the hierarchy and represents the entire system as composition is applied recursively up the hierarchy.

Figure~\ref{sys_model} shows a sample event detector system where automata are depicted as solid rectangles.
The type of the automaton appears in a box in the upper left or lower left corner, e.g., Processor.
The name of a named automaton appears after its type, e.g., Converter C.
Actions are depicted as dotted rectangles and bindings are indicate by dotted lines with an arrow pointing to the input action.
Output actions are suffixed with \emph{!}, e.g., \emph{detect!}, input actions are suffixed with \emph{?}, e.g., \emph{detected?}, and internal actions have neither suffix, e.g., \emph{process1}.
The type associated with a valued action is in parentheses after the name, e.g., \emph{out!(s)}.

The sample event detector system illustrated in Figure~\ref{sys_model} contains three automata (components):  an anonymous Processor, a Converter named C, and a Detector named D.
The Processor automaton is the root and C and D are its children.

\paragraph{Explicit Composition.}
Composition in the I/O automata model is based on actions having the same name and can be applied to an arbitrary number of automata.
Using recursive encapsulation, developers must have the ability to compose automata regardless of the naming convention used by the original developers.
To prepare for dynamic composition, we limit the scope of composition to associating a single output action with a single input action.
Such an association is called a \emph{binding} and consists of a pair $(output, input)$ where $output$ is the name of the output action and $input$ is the name of the input action.

Each automaton contains a set of bindings referring to its own actions, the actions of its children, the actions of its children's children, etc.
The automaton that prescribes a binding is said to \emph{own} that binding.
The global set of bindings can be used to rename input actions before applying composition as defined by the I/O automata model.
A map for renaming inputs is generated by replacing action names with their fully-qualified names in all bindings by tracing from the root automaton.
For example, the fully-qualified name of the $in?(s)$ action of the Converter C in Figure~\ref{sys_model} is $.C.in?(s)$ due to the anonymous root automaton.
Composition as defined by the I/O automata model can be applied after using the map to rename all input actions to their corresponding output action.
The map of bindings for the event detection system is $\{ (.out!(s), .C.in?(s)), (.C.out!(t), .D.in?(t)), (.D.detect!, .detected?) \}$.

\paragraph{Binding rules.}
To be compliant with name-based composition under the I/O automata model, the set of bindings must adhere to certain rules.
First, the output action and input action of a binding must agree on the type being produced and consumed.
For example, an input action consuming an integer value can only be bound to an output action that produces an integer value.
The system shown in Figure~\ref{sys_model} follows this rule since $(out!(s), in?(s))$ agree on type $s$, $(out!(t), in?(t))$ agree on type $t$, and $(detect!, detected?)$ agree on the absence of a value (i.e., a signal).
Second, an input action can be bound to at most one output action.
Composing an input action with multiple output actions would introduce ambiguity into the relationship between the states of the automata and is therefore prohibited by the model.
A binding map where all bindings that mention the same input action are equivalent satisfies this rule.
For dynamics, which are discussed in Section~\ref{dynamics}, we must strengthen this rule to say that an input action can appear at most once in the global binding map.
This strengthened rule means that an input is the sink of at most one binding and that every binding has exactly one owner.
Third, an output action cannot be bound to more than one input action in the same automaton:  the state of the automaton containing the input actions would not be well-defined after the output is executed since the input action effects would be applied in an undefined order.
%% Let $\pi$ be a prefix operator that strips the action name (leaving the path and hence the automaton) of a fully-qualified action name.
%% The following must be true for the global binding map $B$:  $\langle \forall o : (o, z) \in B :: \langle \forall i,j : (o, i) \in B \land (o, j) \in B \land i \neq j:: \pi (i) \neq \pi (j) \rangle \rangle$.
Fourth, an output action cannot be bound to an input action in the same automaton.
Allowing such bindings would result in undefined behavior since the order of the output and input effects is undefined.
The I/O automata model prevents this by requiring that all actions belonging to the same automaton have a unique name.
To elaborate, we usually think about the output action occurring before the input action because a value must be generated.
However, the I/O automata model admits an interpretation where the value is generated first and the effects of the output action and input action are applied non-deterministically.
Graphically, an arrow originating in one component must terminate in another component, even if the latter is a child of the former as Figure~\ref{sys_model} illustrates.
%% To adhere to this rule, the following must be true for the global binding map $B$:  $\langle \forall o, i : (o, i) \in B :: \pi (o) \neq \pi (i) \rangle$.

\paragraph{Concurrent execution.}
Recall that composing I/O automata consists of concatenating state vectors and folding input actions into output actions and that the result of composition is a single equivalent automaton.
Since the state of each automaton is independent, opportunities exist for concurrent execution if the state of each automaton is preserved and a composition is not reduced to a single equivalent automaton.
Each local action implies a set of automata which in turn implies a set of state variables that might be modified.
For an internal action, this set consists of the automaton containing the internal action.
For an output action, this set consists of the automaton containing the output action and the automata that contain the input actions to which the output action is bound.
Two actions can be executed concurrently if their respective sets of state variables are disjoint.
The implied automata sets for each local action in Figure~\ref{sys_model} are: $out!(s) \to \{root, C\}$, $out!(t) \to \{C, D\}$, $process1 \to \{D\}$, $process2 \to \{D\}$, and $detect! \to \{D, root\}$.
Thus, $process1$ and $out!(s)$ can be executed concurrently while $process1$ and $detect!$ can not.

These observations create some interesting opportunities for designing and analyzing concurrent software.
Migrating intense computation to internal actions increases the level of parallelism in a system, i.e., the set of implied automata is small.
A high degree of fan-out (an output action being bound to many input actions) creates a bottle-neck and decreases the level of parallelism, i.e., the set of implied automata is large.
However, the input effects can be applied in parallel since the state of each automaton is independent.
A high degree of fan-in (an automaton has many bound input actions) also creates a bottle-neck by increasing the probability that two sets of implied automata will not be disjoint.
A scheduler might cluster and pin automata to processors to maximize concurrent execution while minimizing inter-processor communication.
Compositions of automata can be statically composed to take advantage of in-lining.
An individual automaton can be decomposed into a number of child automata to increase parallelism.

\paragraph{Parameters and action types.}
Actions in the I/O automata model can be defined using parameters.
Parameters are useful for situations requiring fan-in or session semantics.
For example, consider an automaton with an input action $in?(s)$ that is to be bound to $N$ different output actions.
To get around the limitation that an input can only be bound to one output, we augment the input action with a parameter that indicates the associated output action, e.g., $in?[i](s)$ where $i$ takes on the integer values from 1 to $N$.
\emph{Parameterized} actions take a parameter while \emph{unparameterized} actions do not.
Parameters, for actions that require them, are specified in the binding and serve to identify the action.

The allowed combinations of input, output, and internal actions, unvalued and valued actions, and unparameterized and parameterized actions result in 10 possible action types:
unvalued unparameterized output (uv-up output),
unvalued parameterized output (uv-p output),
valued unparameterized output (v-up output),
valued parameterized output (v-p output),
unvalued unparameterized input (uv-up input),
unvalued parameterized input (uv-p input),
valued unparameterized input (v-up input),
valued parameterized input (v-p input),
unparameterized internal (up internal), and
parameterized internal (p internal).

@@

\paragraph{Example.}
Figure~\ref{example1} depicts a remote sensor system as a composition of automata using a scheme similar to Figure~\ref{sys_model} where parameters are given in brackets after the action name, e.g., \emph{recv?[0](m)}.
The Remote Sensor automaton consists of the Control automaton Z and three Channel automata A, B, and C representing three network connections.
The Control automaton Z consists of the Sensor automaton S, the Filter automaton F, and the Statistics automaton T.
The Control automaton C starts the sampling process with the uv-up output $Z.trigger!$.
The uv-up input $S.trigger?$ is executed atomically with $Z.trigger!$ due to the binding $(Z.trigger!, S.trigger?)$.
When the sampling process is complete, the v-up output $S.sample!(s)$ distributes the sample to the Filter F and Statistics component T via the $(S.sample!(s), F.in?(s))$ and $(S.sample!(s), T.in?(s))$ bindings.
Note that $S.sample!(s)$, $F.in?(s)$, and $T.in?(s)$ all agree on the type of the sample $s$.
The Filter automaton F passes the filtered sample to the Control automaton Z via the $(F.out!(t), Z.sample?(t))$ binding.
The Statistics automaton T calculates the statistics incrementally using the up internal $T.compute$.
The statistics can be polled by the Control automaton Z by issuing a request via the $(Z.request!, T.request?)$ binding and then receiving a response via the $(T.response!(u), Z.response?(u))$ binding.
The Control automaton Z contains a vp-output $Z.send![](m)$ and a vp-input $Z.recv?[](m)$ for sending and receiving network messages using the Channel automata A, B, and C.
The parameters 0, 1, and 2, are used to route messages to Channel automata A, B, and C respectively using the session idea mentioned previously.
Opportunities for concurrent execution exist for the remote sensor depicted in Figure~\ref{example1}.
The set of automata implied by $Z.trigger!$ is $\{Z, S\}$ while the set of automata implied by $T.compute$ is $\{T\}$.
Since the sets are disjoint, the two actions can be executed concurrently.

@@

The I/O automata model assumes that systems are composed of a finite or countably infinite number of automata~\cite{lynch1996distributed}.
Assuming an countably infinite number of participants allows properties of the systems to be stated in terms of an abstract quantity representing the number of members, e.g., $N$, and is reasonable because all real systems are composed of a finite number of members.
The set of interactions in the I/O automata model is also fixed since automata are statically composed.
Dynamic systems are modeled by assuming that only a subset of the all the automata that could possibly exist are actively participating in the system.
To use this technique, a flag is associated with each automaton indicating if it is active or inactive and external actions for ``waking up'' and ``going to sleep'' are defined~\cite{lynch1996distributed}.

While assuming a fixed number of automata is reasonable for modeling, assuming a fixed number of automata for a real system results in over-provisioning or inflexibility.
To develop a component based on the static I/O automata model, a developer must choose a concrete value for the abstract quantity $N$.
System resources are wasted if the actual number of participants is much smaller than $N$.
If the number of participants is greater than $N$, then the system cannot respond to certain situations even though resources might be available.
Consequently, we add the ability to deal with a dynamic set of automata and interactions.

\paragraph{System actions.}
The four indivisible operations for managing a dynamic set of automata are \emph{create}, \emph{bind}, \emph{unbind}, and \emph{destroy}.
Create, bind, unbind, and destroy are collectively called \emph{system actions}.
The \emph{create} operation allows an automaton to create a child automaton.
The \emph{bind} operation allows an automaton to bind an input action to an output action while the \emph{unbind} operation allows an automaton to separate an input action from an output action.
The \emph{destroy} operation allows an automaton to recursively destroy a child automaton and all bindings associated with it.

System actions can fail for a variety of reasons.
A create fails if the automaton that should be added to the set of automata, i.e., the automata being created, already exists.
A bind fails if the automaton associated with the output action doesn't exist, the automaton associated with the input action doesn't exist, or any of the binding rules outlined in Section~\ref{practical} would be violated.
An unbind fails if the binding does not exist.
Note that the owner of binding is taken into account when determining if the binding exists so only the owner of a binding can dissolve it.
A destroy fails if the automaton being destroyed does not exist or is not a child of the automaton requesting the destroy action.

@@

\paragraph{The system automaton.}
The set of automata $A$ and set of bindings $B$ are managed by the \emph{system automaton} which represents the run-time system.
Each automaton is composed (not bound) with the system automaton and has output actions for requesting system actions and input actions for receiving the results of system actions.
The system automaton, in turn, has input actions for receiving requests for system actions, possibly some internal actions for evaluating system action requests, and output actions for responding with the result of system actions.
Figure~\ref{system_action} shows the relationship between a base automaton and the system automaton.

Conceptually, the state variables $A$ and $B$ belong to the system automaton.
However, the state encoded by $A$ and $B$ is shared and used in the execution of every action.
The scheduler uses $A$ to ensure that only actions belonging to automata that exist are selected and uses $B$ to ensure that the appropriate set of input actions receive the value produced when an output action is executed.
Thus, the set of actions implied by any action is the set defined in Section~\ref{practical} \emph{and} and the system automaton.
However, the opportunities for concurrency explored in Section~\ref{practical} are not lost when one considers that the only actions that modify $A$ and $B$ must be serialized, i.e., not executed concurrently with other actions.

@@

To force a dynamic system to behave like a static system, the execution of output actions must be delayed until the set of implied automata matches the set implied by the static system.
To do this, we take the conjunct of the precondition of an output action with a predicate over the set of bindings $B$ called a \emph{binding predicate} to ensure that the output action is not executed until its set of implied automata is correct.
Notice that this only requires reading $B$ and is permitted since all updates to $B$ are serialized as previously discussed.
Thus, a dynamic system can be analyzed as a static system by ignoring system actions and the binding predicates.
A static system can be implemented by adding binding predicates and logic to create the required constellation of automata.
